Applied Measure Theory for Probabilistic Modeling

### 2021-10-01 paper/paper.tex
fix some typos (#162)

\usepackage[normalem]{ulem} %\sout
\usepackage{centernot} %\centernot
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{hyperref}
\setcounter{page}{1}
\include{moritzdefs}
\setlength{\parskip}{0.5em}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\chad}[2][] {\todo[inline,backgroundcolor=orange!20!white, size=\footnotesize,#1]{(Chad) #2 }}
\newcommand{\moritz}[2][] {\todo[inline,backgroundcolor=blue!10!white, size=\footnotesize,#1]{(Moritz) #2 }}
\begin{document}

@@

Let's first consider Bayesian modeling. In the posterior density of the parameter given the observation or data,
\[
p(\theta \mid x) = \frac{p(\theta)\ p(x \mid \theta)}{p(x)}\ ,
\]
the denominator is called the \href{https://en.wikipedia.org/wiki/Marginal_likelihood}{\emph{model evidence}}. Computing it efficiently is often difficult or unfeasible.

@@

Throughout this discussion, it's important to keep in mind that for us the measure-theoretic abstractions are only means to an end.% 
\footnote{The stackoverflow discussion \url{https://mathoverflow.net/q/11591} discusses books on measure theory, M.S.~likes \cite{Shiryaev1996} as starting point and uses \cite{Elstrodt2011} and \cite{Bauer_Heinz1982-01-01} as reference. As we anticipate interest in connecting with the actual theory, we give some pointers to more technical details in the footnotes.}

@@

It has a special role of tying random quantities a program produces and their distributions together, through the notion of random variables, which are functions $\bX$, $\bY$ from $\Omega$ taking values in spaces $X$, $Y$ such as $\{1,2,3,4,5,6\}$ or $\RR^3$.\footnote{Additionally, one requires that random variables are \href{https://en.wikipedia.org/wiki/Measurable_function}{\emph{measurable function}} to be one for which the inverse image of an event $\subset X$ is also an event of $\Omega$.

This is a similar, but much weaker requirement to the definition of a continuous function (``the inverse image of an open set is also open'').}
The package \verb|Omega.jl| \cite{tavares2019language} uses random variables derived from $\Omega$ as  core principle. 
From a computational perspective,
a Julia function taking only an \verb|rng::AbstractRNG| argument, making a number of calls to \verb|rand(rng)|, and returning a value $x\in X$ is a random variable $\bX \colon \Omega\to X$.
A simple example is  \verb|X(rng=Random.GLOBAL_RNG) = rand(rng, 1:6)|.
While mathematically $\bX$ is a function from $\Omega$, the argument $\omega$ is often hidden like the function argument \verb|rng| of its computational counterpart---not a coincidence.

@@

In \verb|MeasureTheory|, measures have abstract super-type `AbstractMeasure`.

@@

There are many cases where it's important for performance to have constant and parameter-dependent terms separated from data-dependent ones. Suppose $\mu$ is a measure with log-density of the form

\[
\log \frac{\dd\mu}{\dd\nu}(x) = \ell(x; \theta) = f(x,\theta) + g(\theta) + C\ .
\]

If we instead observe an iid product of $N$ observations, the log-density is

\[
\begin{aligned}
\sum_{j=1}^N\ell(x_j; \theta) 
    &= \sum_{j=1}^N \left[ f(x_j,\theta) + g(\theta) + C\right] \\
    &= N [g(\theta) + C] + \sum_{j=1}^N  f(x_j,\theta)\ .
\end{aligned}
\]

@@

\verb|MultivariateMeausures.jl|\cite{multivariatemeasures} gives high-performance implementations of \verb|logdensity| for multivariate measures, using \verb|LoopVectorization.jl|\cite{loopvectorization}.

>>>>

\usepackage{inconsolata}
\usepackage[normalem]{ulem} %\sout
\usepackage{centernot} %\centernot
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{hyperref}
\setcounter{page}{1}
\include{moritzdefs}
\setlength{\parskip}{0.5em}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\chad}[2][] {\todo[inline,backgroundcolor=orange!20!white, size=\footnotesize,#1]{(Chad) #2 }}
\newcommand{\moritz}[2][] {\todo[inline,backgroundcolor=blue!10!white, size=\footnotesize,#1]{(Moritz) #2 }}
\begin{document}

@@

Let's first consider Bayesian modeling. In the posterior density of the
parameter $\theta$ given the observation or data $x$,
\[
p(\theta \mid x) = \frac{p(\theta)\ p(x \mid \theta)}{p(x)}\ ,
\]
the denominator is called the \href{https://en.wikipedia.org/wiki/Marginal_likelihood}{\emph{model evidence}}. Computing it efficiently is often difficult or unfeasible.

@@

Throughout this discussion, it's important to keep in mind that for us the measure-theoretic abstractions are only means to an end.% 
\footnote{The stackoverflow discussion \url{https://mathoverflow.net/q/11591} discusses books on measure theory; M.S.~likes \cite{Shiryaev1996} as starting point and uses \cite{Elstrodt2011} and \cite{Bauer_Heinz1982-01-01} as reference. As we anticipate interest in connecting with the actual theory, we give some pointers to more technical details in the footnotes.}

@@

$\Omega$ has a special role of tying random quantities a program produces and their
distributions together, through the notion of random variables, which are
functions $\bX$, $\bY$ from $\Omega$ taking values in spaces $X$, $Y$ such as
$\{1,2,3,4,5,6\}$ or $\RR^3$.\footnote{Additionally, one requires a random
variable to be a \href{https://en.wikipedia.org/wiki/Measurable_function}{\emph{measurable
function}}, that is, one for which the inverse image of an event $\subset X$ is
also an event of $\Omega$.
This is a similar, but much weaker requirement to the definition of a continuous function (``the inverse image of an open set is also open'').}
The package \verb|Omega.jl| \cite{tavares2019language} uses random variables
derived from $\Omega$ as  core principle.

From a computational perspective,
a Julia function taking only an \verb|rng::AbstractRNG| argument, making a number of calls to \verb|rand(rng)|, and returning a value $x\in X$ is a random variable $\bX \colon \Omega\to X$.
A simple example is  \verb|X(rng=Random.GLOBAL_RNG) = rand(rng, 1:6)|.
While mathematically $\bX$ is a function from $\Omega$, the argument $\omega$ is often hidden like the function argument \verb|rng| of its computational counterpart---not a coincidence.

@@

In \verb|MeasureTheory|, measures have abstract super-type \verb|AbstractMeasure|.

@@

There are many cases where it's important for performance to have constant and parameter-dependent terms separated from data-dependent ones. Suppose $\mu$ is a measure with log-density of the form
\[
\log \frac{\dd\mu}{\dd\nu}(x) = \ell(x; \theta) = f(x,\theta) + g(\theta) + C\ .
\]

If we instead observe an iid product of $N$ observations, the log-density is
\[
\begin{aligned}
\sum_{j=1}^N\ell(x_j; \theta) 
    &= \sum_{j=1}^N \left[ f(x_j,\theta) + g(\theta) + C\right] \\
    &= N [g(\theta) + C] + \sum_{j=1}^N  f(x_j,\theta)\ .
\end{aligned}
\]

@@

\verb|MultivariateMeasures.jl|\cite{multivariatemeasures} gives high-performance implementations of \verb|logdensity| for multivariate measures, using \verb|LoopVectorization.jl|\cite{loopvectorization}.
